\ttitle

### 2015-08-18 Chapters/BodyComp.tex
Body Composition - Nearly there
- Almost done
- Just need to check minor things
- LEAVE THIS BE NOW

The results of this study show that the most important cardiopulmonary exercise test parameters as used for preoperative risk evaluation in surgery are influenced significantly by the patient's body composition.

@@

The positive correlation between absolute $\dot{V}_{O_2}$ and skeletal muscle area is easily explained by the physiology of aerobic exercise. During periods of increased physical activity, the greater oxygen demand is primarily due to increased metabolic activity within the skeletal muscle.

Current convention is to report $\dot{V}_{O_2}$ measured at cardiopulmonary exercise testing according the following formula: 
\[Corrected\ \dot{V}_{O_2} (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ \dot{V}_{O_2}\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]

In a previous analysis (refer to chapter and table), we reported that there was a significant negative correlation between $\dot{V}_{O_2}$ at anaerobic threshold and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease.

The results of the present study suggest that the negative correlation between corrected $VO_{2}$ (mls/kg/min) and BMI is consequent to the reporting convention rather than due to any pathophysiological effect of obesity.

The loss of the strong positive correlation between absolute $VO_{2}$ (litres/min) and skeletal muscle area after correcting for body weight further supports the argument that the corrected value under-reports aerobic capacity in obese patients. Moreover, the lack of correlation between pulmonary function tests, tidal volume and minute ventilation and adipose tissue area as well as the slight but statistically significant positive correlation between O2Pulse and adipose tissue area appear to suggest that adiposity did not contribute to poor cardiopulmonary exercise performance in this cohort of patients.

@@

Seltzer reported in his 1940 study of 34 subjects, that the individuals who were more "lateral" than "linear" had lower oxygen intakes per kilo body weight

Batterham et al studied 1314 apparently healthy men employed at the National Aeronautics and Space Administration Johnson Space Center in Houston, Texas \parencite{batterham_modeling_1999}. The authors report that as body mass increased, the proportion composed of fat-free mass decreased. They also found that fat-free mass had a linear relationship with oxygen consumption while total body mass did not. They suggest that ideally estimates of fat-free mass should be used in the representation of oxygen consumption to allow more reliable comparison between subjects.

Janz et al studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998 \parencite{janz_longitudinal_1998}. Aerobic capacity in the form of $\dot{V}_{O_2}$peak was evaluated annually in 126 children (mean age 10.3 years) for five years. Body composition changes were also tracked over this period. They reported on the changes in body composition that occur over time and the differences in these changes between circum-pubertal boys and girls. They reported on the significant difficulties in normalising $\dot{V}_{O_2}$ using total body mass and suggested that fat-free mass was the most appropriate variable for normalising $\dot{V}_{O_2}$. They found that $\dot{V}_{O_2}$ normalised using total body mass underestimated aerobic fitness levels of heavier boys and girls. However, this underestimation was greater in girls than in boys.

Goran et al reported that total body fat did not affect maximal aerobic capacity \parencite{goran_total_2000}. They reported on $\dot{V}_{O_2}$max in obese women before and after weight loss. $\dot{V}_{O_2}$max corrected for total body weight was significantly lower in the obese state while $\dot{V}_{O_2}$max corrected for fat-free mass did not change significantly after weight loss. They also reported that the limiting factor in the obese state was not the cardio-respiratory system but the fact that it was more difficult for obese individuals to do the same amount of work as a normal weight person in weight-bearing activities. This is likely due to the extra fat mass in these individuals that did not contribute to aerobic capacity but instead may increase the exercise load.

These findings have been replicated by several other authors in different subject groups \parencite{loftin_scaling_2001,  lemaitre_maximum_2006,savonen_current_2012, krachler_cardiopulmonary_2014}. Several of the above studies also recommend using allometric scaling to avoid the confounding effects of total body weight. However, this has not gained widespread clinical use.

In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in a two groups of physically fit men categorised based on their body fat percentage \parencite{maciejczyk_influence_2014}. They reported that high body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat. However, when $\dot{V}_{O_2}$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups. They, similar to Goran et al\parencite{goran_total_2000}, recommend that $\dot{V}_{O_2}$ be normalised to lean body mass rather than total body weight.

@@

\textit{`Spurious correlation'} is a tern introduced by English mathematician and biometrician, Karl Pearson in 1896. He used the term to describe correlations that occurred as a result of using ratios instead of absolute values rather than because of any actual correlations between the variables being studied\parencite{pearson_mathematical_1896}
Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}. While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome. However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure. This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}. The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}.

Swart and Carlisle reported that $\dot{V}_{O_2}$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes \parencite{swart_case-controlled_2012}. However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT groups (24\% vs 51\%). The average $\dot{V}_{O_2}$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min. In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low $\dot{V}_{O_2}$AT group was 51\% while it was 28\% in the group with normal AT \parencite{wilson_impaired_2010}. There was no data presented on body mass index in this study.

This is similar to the findings in our cohort of patients. This may have been due to the increased incidence of obesity especially in the subcutaneous plane as we have found in our cohort of patients as shown in Fig. \ref{fig:bc_figure1}.

It is clear from the review presented in Chapter 1, that cardiopulmonary exercise testing is useful in predicting risk after major surgery. Cardiopulmonary exercise testing has become ubiquitous in the preoperative workup of complex surgical patients. However, the results of the present study suggest that the results especially in the obese, female patient must be interpreted with caution, especially when used to select patients who may be declined surgery based on their cardiopulmonary exercise test results.

\subsection{Measuring impact of Prehabilitation}

Where time to surgery is not critical, prehabilitation has gained an increasingly important role in optimising patients for surgery and mitigating the effects of neoadjuvant oncological therapy. Cardiopulmonary exercise testing has been reported to be a useful objective measure of the impact of prehabilitation in surgical patients \parencite{west_effect_2015}.

The design of such prehabilitation programs must not depend solely on body weight adjusted parameters of cardiopulmonary exercise testing when assessing the success of the interventions in these programs. Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with other parameters that are not affected by body composition such as O2Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity.

>>>>

The results of this study demonstrate that the most important cardiopulmonary exercise test parameters used for preoperative risk evaluation in surgery are influenced significantly by the patient's body composition. Both $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak are significantly affected by patient's body habitus when scaled using total body weight. 
Furthermore, the results suggest that there may be alternate methods of scaling of $\dot{V}_{O_2}$ that removes the effect of body composition on these parameters. 
The results also demonstrate the disproportionate decrease in skeletal muscle area with associated increased in adipose tissue with increasing body mass index. 
Taken together, these results would suggest that caution should be exercised in interpreting $\dot{V}_{O_2}$ (corrected for total body weight, mls/kg/min) in patients with a raised BMI as abnormally low values in these patients may not necessarily be due to poor aerobic fitness or lack of cardiopulmonary reserve.

@@

Increased physical activity such as during cardiopulmonary exercise testing results in greater oxygen consumption ($\dot{V}_{O_2}$). The increased demand for oxygen is primarily due to increased metabolic activity within the exercising skeletal muscle.  Therefore, the positive correlation between absolute $\dot{V}_{O_2}$ and skeletal muscle area as has been found here is easily explained by the physiology of aerobic exercise.

However, current convention is to report $\dot{V}_{O_2}$ measured at cardiopulmonary exercise testing according to the following formula: 
\[Corrected\ \dot{V}_{O_2} (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ \dot{V}_{O_2}\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]

In chapter \ref{ch_cpet_jaundice}, we reported that there was a significant negative correlation between corrected $\dot{V}_{O_2}$ (ml/kg/min) at the anaerobic threshold as well as at peak exercise and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease (Tables \ref{table:cpet_oj_at_regression} and \ref{table:cpet_oj_peak_regression}). The results of the present study demonstrate that this negative correlation is consequent to the scaling convention used rather than due to any pathophysiological effect of obesity.

The loss of the strong positive correlation between absolute $VO_{2}$ (litres/min) and skeletal muscle area after correcting for total body weight further supports the argument that the corrected value for $\dot{V}_{O_2}$ under-reports aerobic capacity in obese patients. Moreover, the lack of correlation between pulmonary function tests, tidal volume and minute ventilation and adipose tissue area as well as the slight but statistically significant positive correlation between adipose tissue and $O_2$Pulse at anaerobic threshold as well as peak exercise but not at rest appear to suggest that adiposity did not contribute to poor cardiopulmonary exercise performance in this cohort of patients.

@@

Seltzer reported in his 1940 study of 34 subjects, that individuals who were more `lateral' than `linear' had lower oxygen intakes per kilo body weight.\parencite{seltzer_body_1940} Tanner in his article titled \textit{`Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation'}\parencite{tanner_fallacy_1949} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass. In a detailed analysis comparing $\dot{V}_{O_2}$ and body build, he concluded that \textit{'as the index wt./stature increases, O2/wt. must be expected to decrease purely as a result of the method used for representing the data.'}

Batterham et al studied 1314 apparently healthy men employed at the National Aeronautics and Space Administration Johnson Space Center in Houston, Texas \parencite{batterham_modeling_1999}. The authors reported that as body mass increased, the proportion composed of fat-free mass decreased. They also found that fat-free mass had a linear relationship with oxygen consumption while total body mass did not. They suggest that ideally estimates of fat-free mass should be used in the representation of oxygen consumption to allow more reliable comparison between subjects. These findings are similar to our results where the proportion of skeletal muscle area decreasing significantly as body mass index increased with a greater decrease in females than males. Moreover, the linear relationship between skeletal muscle area and absolute $\dot{V}_{O_2}$ has also been clearly demonstrated in our findings.

We used the Boer formula \parencite{boer_estimated_1984}to estimate lean body mass and used this to normalise $\dot{V}_{O_2}$. The resulting value for $\dot{V}_{O_2}$ showed no `spurious' correlations with body composition, height, total body weight or body mass index. These findings are similar to the recommendations made by Janz et al who studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998 \parencite{janz_longitudinal_1998}. Aerobic capacity in the form of $\dot{V}_{O_2}$peak was evaluated annually in 126 children (mean age 10.3 years) for five years. Body composition changes were also tracked over this period. They reported on the changes in body composition that occur over time and the differences in these changes between circum-pubertal boys and girls. They reported on the significant difficulties in normalising $\dot{V}_{O_2}$ using total body mass and suggested that fat-free mass was the most appropriate variable for normalising $\dot{V}_{O_2}$. They found that $\dot{V}_{O_2}$ normalised using total body mass underestimated aerobic fitness levels of heavier boys and girls. However, this underestimation was greater in girls than in boys.

Our results demonstrate that there is no correlation between absolute oxygen consumption at rest or during exercise and the amount of adipose tissue. This further emphasises the fact that adipose tissue is metabolically inactive during exercise although it contributes to increased work load. Therefore, dividing absolute $\dot{V}_{O_2}$ by total body weight in overweight/obese subjects whose extra weight is largely due to adipose tissue results in lower corrected $\dot{V}_{O_2}$ as we have demonstrated here. 
Goran et al reported that total body fat did not affect maximal aerobic capacity \parencite{goran_total_2000}. They reported on $\dot{V}_{O_2}$max in obese women before and after weight loss. $\dot{V}_{O_2}$max corrected for total body weight was significantly lower in the obese state while $\dot{V}_{O_2}$max corrected for fat-free mass did not change significantly after weight loss. They also reported that the limiting factor in the obese state was not the cardio-respiratory system but the fact that it was more difficult for obese individuals to do the same amount of work as a normal weight person in weight-bearing activities. This is likely due to the extra fat mass in these individuals that did not contribute to aerobic capacity but instead may increase the exercise load.

These findings have been replicated by several other authors in different subject groups \parencite{loftin_scaling_2001,  lemaitre_maximum_2006,savonen_current_2012, krachler_cardiopulmonary_2014}. Several of the above studies also recommended using allometric scaling to avoid the confounding effects of total body weight. However, this has not gained widespread clinical use.

In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in two groups of physically fit men categorised based on their body fat percentage \parencite{maciejczyk_influence_2014}. They reported that high body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat. However, when $\dot{V}_{O_2}$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups. They, similar to Goran et al\parencite{goran_total_2000}, recommend that $\dot{V}_{O_2}$ be normalised to lean body mass rather than total body weight.

@@

The term \textit{`spurious correlation'} was first introduced by English mathematician and biometrician, Karl Pearson in 1896. He used the term to describe correlations that occurred as a result of using ratios instead of absolute values rather than because of any actual correlations between the variables being studied \parencite{pearson_mathematical_1896}.

Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}. While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome. However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tool concluded that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure. This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}. The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}.

Swart and Carlisle reported that $\dot{V}_{O_2}$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes \parencite{swart_case-controlled_2012}. However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT groups (24\% vs 51\%). The average corrected $\dot{V}_{O_2}$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min. In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low $\dot{V}_{O_2}$AT group was 51\% while it was 28\% in the group with normal AT \parencite{wilson_impaired_2010}. There was no data presented on body mass index in this study. We reported in Chapter \ref{ch_cpet_jaundice} that females were more likely to have a lower $\dot{V}_{O_2}$AT (Table \ref{table:cpet_oj_at_regression}) and lower $\dot{V}_{O_2}$Peak (Table \ref{table:cpet_oj_peak_regression}). However, all the results in the present study were controlled for the effect of sex.

The `obesity paradox' would suggest that, in fact, obese patients are not necessarily at an increased risk of postoperative complications. Several authors have reported that obesity is not a risk factor for major complications after pancreaticoduodenectomy \parencite{khan_does_2010, tsai_impact_2010, balentine_obesity_2011}. 
In a large study of 118,707 patients undergoing non-bariatric general surgery, the authors reported that the risk of death was `paradoxically' low in the overweight and obese group of patients \parencite{mullen_obesity_2009}. The results of the present study show that $\dot{V}_{O_2}$ as is currently reported for clinical use will also be low in this very same cohort of patients. 
This obesity paradox is particularly apparent in patients with heart failure where in spite of a low $\dot{V}_{O_2}$ (corrected for total body weight), obese patients have better survival rates. Horwich and co-workers reported in a study of 2331 patients that body mass index was associated with a low $\dot{V}_{O_2}$Peak (ml/kg/min) independent of all other factors including age, diabetes, cardiac disease, New York Heart Association Class of cardiac failure and race \parencite{horwich_relationship_2009}.

It is clear from the review presented in Chapter \ref{ch_intro} as well as the results presented in Chapter \ref{ch_cpet_outcomes}, that cardiopulmonary exercise testing is useful in predicting risk after major surgery. Cardiopulmonary exercise testing has become ubiquitous in the preoperative assessment of complex surgical patients. However, the present study demonstrates that cardiopulmonary exercise test parameters in the overweight/obese patient must be interpreted with caution, especially when used to select patients who may be declined surgery based on these results.

\subsection{Measuring impact of prehabilitation}

Where time to surgery is not critical, prehabilitation has gained an increasingly important role in optimising patients for surgery, reducing complications \parencite{jones_effects_2007, dronkers_preoperative_2010, valkenet_effects_2010, gillis_prehabilitation_2014} and mitigating the effects of neoadjuvant oncological therapy\parencite{west_effect_2015}.
Cardiopulmonary exercise testing has been reported to be a useful objective measure of the impact of prehabilitation in surgical patients \parencite{west_effect_2015}.

The success of such prehabilitation programs must not be determined solely by $\dot{V}_{O_2}$ corrected for total body weight. Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with improvement in other parameters that are not affected by body composition such as $O^2$Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity. The combination of nutritional assessment and advice, home-based exercise programs and weight loss will help improve aerobic capacity and eventually postoperative outcomes.

Further studies are required to validate the findings presented here as well as to identify optimal scaling methods for cardiopulmonary exercise test parameters. This will improve the clinical utility of cardiopulmonary exercise testing in the preoperative risk assessment as well as optimisation of patients undergoing major surgery.
